---
title: "Data Visualization Lab"
subtitle: "Creating Effective Visual Stories with Data"
author: "Rod Albuyeh"
date: today
format: pdf
execute:
  echo: true
  eval: true
  warning: false
  error: false
  cache: true
  tidy: true
editor: 
  markdown: 
    wrap: 72
---

# Introduction

Data visualization is one of the most powerful tools in a researcher's toolkit. Good visualizations can reveal patterns, communicate findings clearly, and tell compelling stories with data. Poor visualizations can mislead, confuse, or obscure important insights.

## Learning Objectives

By the end of this lab, you will be able to:

1. **Understand the grammar of graphics** - the foundational principles behind ggplot2
2. **Create effective univariate visualizations** to explore single variables
3. **Build informative bivariate plots** to examine relationships between variables
4. **Design multivariate visualizations** using color, size, and faceting
5. **Apply design principles** to create clear, accessible, and professional plots
6. **Avoid common visualization pitfalls** that can mislead or confuse audiences

## The Grammar of Graphics

The `ggplot2` package is built on the "grammar of graphics" - a systematic approach to creating visualizations. Think of it like constructing a sentence: you need subjects, verbs, and objects arranged in a logical structure.

Every ggplot has these essential components:

- **Data**: Your dataset (the subject)
- **Aesthetics (aes)**: How variables map to visual properties (the relationships)
- **Geometries (geoms)**: The type of visual representation (the action)

Optional components add sophistication:

- **Scales**: Control how data values map to visual values
- **Facets**: Create subplots for different groups
- **Themes**: Control the overall appearance
- **Labels**: Add titles, captions, and axis labels

# Setup and Data Exploration

```{r message = FALSE}
# Essential libraries for data visualization and manipulation
library(tidyverse)    # Includes ggplot2, dplyr, and other useful packages
library(poliscidata)  # Contains political science datasets

# Optional: Install if not already available
# install.packages(c("tidyverse", "poliscidata", "ggthemes"))
```

## Understanding Our Dataset

Before creating any visualization, we must understand our data. Let's explore the `world` dataset from the `poliscidata` package.

```{r}
# Get information about the dataset
?world  # Opens help documentation

# View the structure of the data
glimpse(world)  # Shows variable types and first few values
```

```{r}
# Get basic summary statistics
summary(world)
```

```{r echo=FALSE}
# Check the dimensions and variable names
cat("Dataset dimensions:", dim(world)[1], "countries,", dim(world)[2], "variables\n")
```

**Key variables we'll explore:**

- `gdppcap08`: GDP per capita in 2008
- `lifeex_total`: Life expectancy 
- `pop_total`: Total population
- `regionun`: UN regional classification

**Key Point**: Always explore your data before visualizing. Understanding the scale, distribution, and missing values of your variables will inform your visualization choices.

# Visualization Fundamentals

## The ggplot2 Approach

The `ggplot2` package follows a layered approach to building visualizations. Think of it as painting: you start with a canvas (data), add a base layer (aesthetics), then build up with additional layers (geometries, themes, labels).

### Basic Structure
```
ggplot(data = your_data) +           # Canvas + paint
  aes(x = variable1, y = variable2) +  # What goes where
  geom_point() +                       # How to display it
  labs(title = "Your Title")           # Labels and annotations
```

## Data Visualization Best Practices

Before we start creating plots, let's establish some principles for effective visualization:

### 1. **Choose the Right Chart Type**
- **Continuous variables**: histograms, density plots, box plots
- **Categorical variables**: bar charts, dot plots
- **Relationships**: scatter plots, line graphs
- **Distributions by group**: faceted plots, grouped charts

### 2. **Design for Your Audience**
- Use clear, descriptive titles and axis labels
- Choose colors that are accessible (avoid red-green combinations)
- Ensure text is large enough to read
- Remove unnecessary elements (chart junk)

### 3. **Tell a Story**
- Every visualization should have a clear message
- Guide the reader's eye to the most important information
- Use annotations to highlight key findings

Visit the [ggplot2 reference](https://ggplot2.tidyverse.org/reference/) for comprehensive documentation.

# Univariate Plots: Exploring Single Variables

Univariate plots help us understand the distribution of a single variable. Different variable types require different approaches.

## Continuous Variables

For continuous variables like GDP per capita, we want to see the shape of the distribution, central tendency, and spread.

```{r}
# Creating a histogram of GDP per capita
# Histograms show the frequency distribution of continuous data

ggplot(data = world) +
  aes(x = gdppcap08) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  labs(
    title = "Distribution of GDP per Capita (2008)",
    x = "GDP per Capita (USD)",
    y = "Number of Countries",
    caption = "Source: World Bank via poliscidata package"
  ) +
  theme_minimal()
```

**Interpretation**: Notice the right-skewed distribution - most countries have lower GDP per capita, with a few wealthy outliers.

```{r}
# Box plots reveal different information about the same variable
# They show median, quartiles, and outliers clearly

ggplot(data = world) +
  aes(y = gdppcap08) +  # Note: y aesthetic for vertical box plot
  geom_boxplot(fill = "lightcoral", alpha = 0.7) +
  labs(
    title = "GDP per Capita Distribution (Box Plot)",
    y = "GDP per Capita (USD)",
    caption = "Box shows median and quartiles; dots show outliers"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),  # Remove x-axis text
        axis.ticks.x = element_blank()) # Remove x-axis ticks
```

## Comparing Different Approaches

Let's see how different visualizations reveal different aspects of the same data:

```{r}
# Density plot - smooth version of histogram
ggplot(data = world) +
  aes(x = gdppcap08) +
  geom_density(fill = "purple", alpha = 0.5) +
  labs(
    title = "GDP per Capita - Density Plot",
    x = "GDP per Capita (USD)",
    y = "Density"
  ) +
  theme_minimal()
```

### 🎯 **Exercise 1: Your Turn** (5 minutes)

Choose a different continuous variable from the `world` dataset and create **two different** univariate visualizations. Consider these options:

- `lifeex_total` (life expectancy)
- `pop_total` (population) 
- `gdp08` (total GDP)

**Available geoms**: `geom_histogram()`, `geom_density()`, `geom_boxplot()`, `geom_violin()`

**Challenge**: Add appropriate titles, labels, and colors to make your plots publication-ready.

```{r}
# Your code here - Visualization 1


```

```{r}
# Your code here - Visualization 2


```

**Reflection Questions**: 

1. What does each visualization tell you about your variable's distribution?
2. Which visualization do you find more informative and why?

# Bivariate Plots: Exploring Relationships

Bivariate plots help us understand relationships between two variables. The choice of visualization depends on the types of variables involved.

## Continuous vs. Continuous: Scatter Plots

Scatter plots are ideal for examining relationships between two continuous variables.

```{r}
# Creating a scatter plot to examine the relationship between wealth and health
# This is a classic relationship in development economics

p2 <- ggplot(world) +
  aes(x = gdppcap08, y = lifeex_total) + 
  geom_point(alpha = 0.6, size = 2) +
  labs(
    title = "The Wealth-Health Relationship",
    x = "GDP per Capita (USD, 2008)",
    y = "Life Expectancy (years)",
    caption = "Each point represents one country"
  ) +
  theme_minimal()

p2
```

**What do you see?** There appears to be a positive relationship - wealthier countries tend to have higher life expectancy.

## Adding Trend Lines

Trend lines help us see patterns more clearly and can reveal the strength of relationships.

```{r}
# Adding a fitted line to show the overall trend
# The gray band shows the confidence interval

p2 +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(
    title = "Wealth-Health Relationship with Linear Trend",
    subtitle = "Gray band shows 95% confidence interval"
  )
```

```{r}
# Sometimes a curved line fits better than a straight line
# method = "loess" creates a flexible, curved trend line

p2 +
  geom_smooth(method = "loess", se = TRUE, color = "blue") +
  labs(
    title = "Wealth-Health Relationship with Curved Trend",
    subtitle = "Curved line may better capture the relationship"
  )
```

**Key Insight**: The relationship might not be perfectly linear - there may be diminishing returns to wealth in terms of health outcomes.

# Advanced Customization: Making Professional Plots

Good data visualization is as much about design as it is about data. Let's learn how to create publication-quality plots.

## Color and Aesthetics

Colors should serve a purpose - they can highlight important information, group related data, or simply make plots more appealing.

```{r}
# Using color strategically to enhance the message
# Here, we use a professional color palette

p2 +
  geom_point(size = 2.5, alpha = 0.7, color = "#2E86AB") +  # Professional blue
  geom_smooth(method = "lm", 
              color = "#A23B72",      # Complementary color for trend line
              fill = "#F18F01",       # Warm color for confidence band
              alpha = 0.3) +
  labs(
    title = "Economic Development and Life Expectancy",
    subtitle = "Higher GDP per capita is associated with longer life expectancy",
    x = "GDP per Capita (USD, 2008)",
    y = "Life Expectancy (years)",
    caption = "Source: World Bank data via poliscidata package"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11, color = "gray40"),
    axis.title = element_text(size = 10),
    plot.caption = element_text(size = 8, color = "gray50")
  )
```

## The Power of Good Labels

Clear, informative labels are crucial for effective communication.

```{r}
# Example of comprehensive labeling for a professional report
ggplot(world) +
  aes(x = gdppcap08, y = lifeex_total) +
  geom_point(alpha = 0.6, size = 2, color = "steelblue") +
  geom_smooth(method = "loess", se = FALSE, color = "darkred", size = 1.2) +  # Changed to loess for curved relationship
  scale_x_continuous(
    labels = scales::dollar_format(),  # Format x-axis as currency
    breaks = seq(0, 80000, 20000)     # Custom break points
  ) +
  scale_y_continuous(
    breaks = seq(40, 85, 10)          # Custom y-axis breaks
  ) +
  labs(
    title = "Wealth and Health Across Nations",
    subtitle = "Higher GDP per capita associated with longer life expectancy",
    x = "GDP per Capita (USD)",
    y = "Life Expectancy (years)",
    caption = "Data: World Bank | Each point represents one country"
  ) +
  theme_classic() +  # Clean, publication-ready theme
  theme(
    plot.title = element_text(size = 16, face = "bold", margin = margin(b = 10)),
    plot.subtitle = element_text(size = 12, color = "gray30", margin = margin(b = 15)),
    axis.title = element_text(size = 11),
    plot.caption = element_text(size = 9, color = "gray50", hjust = 0)
  )
```

**Design Principles Applied:**

- **Hierarchy**: Title is largest, subtitle smaller, caption smallest
- **Color**: Limited palette with purposeful color choices
- **Spacing**: Appropriate margins and spacing between elements
- **Clarity**: Professional formatting and clear labels

# Working with Scales and Transformations

Choosing appropriate scales is crucial for accurate data representation. Sometimes transforming your data can reveal patterns that aren't visible in the original scale.

## Axis Limits: When and Why to Use Them

```{r}
# Should we start axes at zero? It depends on your message and data type
# For life expectancy, starting at 0 might be misleading

# Starting y-axis at 0 - compresses the visible differences
p2 +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  ylim(0, 100) +
  labs(
    title = "Starting Y-axis at Zero",
    subtitle = "Differences between countries appear smaller"
  )
```

```{r}
# Focusing on the actual data range - highlights real differences
p2 +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  ylim(40, 85) +
  labs(
    title = "Focusing on Data Range",
    subtitle = "Real differences between countries are more visible"
  )
```

**When to start at zero**: For quantities where zero is meaningful (counts, percentages, absolute amounts)
**When NOT to start at zero**: For measurements where zero isn't meaningful or relevant (temperature, test scores, life expectancy)

## Log Transformations: Revealing Hidden Patterns

Sometimes the relationship between variables is better understood on a logarithmic scale. **But this isn't just mathematical trickery** - there are real-world reasons why log scales make sense.

### The Intuition: It's About Proportional Changes

Think about it this way: What matters more for quality of life - going from $1,000 to $2,000 income per year, or going from $50,000 to $51,000? 

- **$1,000 → $2,000**: You doubled your income! Huge life change.
- **$50,000 → $51,000**: A nice raise, but not life-changing.

Both are $1,000 increases, but the *proportional* change is what matters for things like health outcomes. Log scales capture these proportional relationships.

```{r}
# The original relationship on linear scales
# Here, small countries and large countries seem to follow different patterns
p2 +
  geom_point(alpha = 0.6, color = "darkblue") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(
    title = "Linear Scale: Hard to See the True Relationship",
    subtitle = "Rich countries seem to cluster at the right, making the pattern unclear"
  )
```

```{r}
# Using log10 transformation for GDP per capita
# Now we can see that the relationship is actually quite consistent across income levels
p2 +
  geom_point(alpha = 0.6, color = "darkgreen") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  scale_x_log10(labels = scales::dollar_format()) +
  labs(
    title = "Log Scale: The True Pattern Emerges",
    subtitle = "Doubling income has similar health effects whether you're poor or rich",
    x = "GDP per Capita (USD, log scale)"
  )
```

### Why This Makes Sense in the Real World

**Log transformations are appropriate when:**

- **Proportional changes matter more than absolute changes** (like income → health)
- **The effect of "doubling" something is consistent** across different starting points
- **You have data spanning many orders of magnitude** (from $500 to $50,000+ per capita)

**Real examples where log scales reveal truth:**

- **Population growth**: Going from 1M to 2M people vs 10M to 11M people
- **Sound**: Decibels are already on a log scale (whisper vs rock concert)
- **Earthquakes**: Richter scale is logarithmic (magnitude 6 vs 7 is 10x stronger)
- **Economic development**: Doubling GDP has similar effects on quality of life regardless of starting point

**This isn't data manipulation** - it's revealing the underlying relationship that linear scales obscure!

### 🎯 **Exercise 2: Scale Transformations** (3 minutes)

Look back at your univariate or bivariate plots from Exercise 1. Consider:

1. **For univariate plots**: Would a log transformation help if your variable is highly skewed?
2. **For bivariate plots**: Does the relationship look curved? Would log-transforming one or both variables help?

Try applying a transformation and compare the results.

```{r}
# Your transformation experiment here


```

**Helpful resource**: [ggplot2 axis scales guide](http://www.sthda.com/english/wiki/ggplot2-axis-scales-and-transformations)

# Multivariate Visualizations: Adding More Dimensions

Real-world relationships often involve multiple variables. We can incorporate additional variables using aesthetics like size, color, and faceting.

## Adding a Third Variable with Size

Let's add population as a third dimension to our wealth-health relationship.

```{r}
# Adding population as the size of points
# This creates a "bubble chart" - a scatter plot where point size represents a third variable

p3 <- ggplot(world) +
  aes(x = gdppcap08, y = lifeex_total, size = pop_total) +
  geom_point(alpha = 0.6, color = "steelblue") +
  scale_size_continuous(
    name = "Population",
    range = c(1, 12),  # Control the range of point sizes
    guide = "none"     # Remove the legend entirely - size is intuitive
  ) +
  labs(
    title = "Wealth, Health, and Population Size",
    subtitle = "Point size represents population (larger = more people)",
    x = "GDP per Capita (USD)",
    y = "Life Expectancy (years)",
    caption = "Source: World Bank data"
  ) +
  theme_minimal() +
  theme(legend.position = "right")

p3
```

**What stories emerge?** 

- Large countries have lower GDP per capita but vary in life expectancy
- Small, wealthy countries often have high life expectancy
- Some large countries are both wealthy and have high life expectancy

## Faceting: Small Multiples

Faceting creates separate panels for different groups, allowing us to see patterns within subgroups.

```{r}
# Creating separate plots for each UN region
# This helps us see if the wealth-health relationship varies by region

p3 + 
  facet_wrap(~regionun, scales = "free") +  # scales = "free" allows different axis ranges
  labs(
    title = "Wealth-Health Relationship by World Region",
    subtitle = "Each panel shows countries from one UN region"
  ) +
  theme(
    strip.text = element_text(size = 8),  # Smaller text for facet labels
    axis.text = element_text(size = 7)    # Smaller axis text
  )
```

**Regional Insights:**

- Does the wealth-health relationship hold across all regions?
- Which regions show the strongest/weakest relationships?
- Are there regional outliers that don't follow the global pattern?

### 🎯 **Exercise 3: Multivariate Exploration** (5 minutes)

Take one of your previous plots and add a third variable using either:

1. **Size aesthetic**: `aes(size = variable_name)`
2. **Color aesthetic**: `aes(color = variable_name)` 
3. **Faceting**: `facet_wrap(~variable_name)`

Consider what story the additional variable tells about your data.

```{r}
# Your multivariate visualization here


```

# Professional Themes and Styling

The right theme can make your visualization look polished and professional. ggplot2 offers many built-in themes, and additional packages provide even more options.

```{r}
# Load additional theme options
library(ggthemes)  # install.packages("ggthemes") if needed

# Create a base plot to demonstrate different themes
base_plot <- ggplot(world) +
  aes(x = gdppcap08, y = lifeex_total) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, color = "darkred") +
  labs(
    title = "Economic Development and Life Expectancy",
    x = "GDP per Capita (USD)",
    y = "Life Expectancy (years)"
  )
```

```{r}
# Built-in ggplot2 themes
base_plot + theme_minimal() + labs(subtitle = "theme_minimal() - clean, modern")
base_plot + theme_classic() + labs(subtitle = "theme_classic() - traditional, publication-ready")
base_plot + theme_void() + labs(subtitle = "theme_void() - minimalist, focus on data")
```

```{r}
# Themes from ggthemes package
base_plot + theme_economist() + labs(subtitle = "theme_economist() - Economist magazine style")
base_plot + theme_fivethirtyeight() + labs(subtitle = "theme_fivethirtyeight() - FiveThirtyEight website style")
base_plot + theme_tufte() + labs(subtitle = "theme_tufte() - Edward Tufte's minimalist approach")
```

**Choosing Themes:**

- **Academic papers**: `theme_classic()` or `theme_minimal()`
- **Business presentations**: `theme_economist()` or custom corporate themes
- **Web/social media**: `theme_fivethirtyeight()` or `theme_modern()`
- **Focus on data**: `theme_void()` or `theme_tufte()`

# Lab Assignment: Creating Publication-Quality Visualizations

## Overview

Your task is to create two high-quality visualizations using the `world` dataset that demonstrate your understanding of effective data visualization principles. Choose variables **different** from those used in the lab examples (avoid `gdppcap08` and `lifeex_total` as your main variables).

## Requirements

### Part 1: Univariate Visualization (5 points)

Create **one univariate plot** that explores the distribution of a single variable.

**Technical Requirements:**

- Choose an appropriate geom for your variable type
- Apply proper styling (colors, transparency, etc.)
- Include meaningful axis labels and title
- Use a professional theme
- Hide code in final output (`echo = FALSE`)

**Variables to consider:** `pop_total`, `gdp08`, `literacy`, `gini`, `dem_score14`, `civlib_score14`

### Part 2: Bivariate Visualization (5 points)

Create **one bivariate plot** that examines the relationship between two variables.

**Technical Requirements:**
- Use appropriate geom(s) for your variable types
- Add a trend line if appropriate
- Consider scale transformations if needed
- Apply professional styling and color choices
- Include comprehensive labels (title, subtitle, axis labels, caption)
- Use a professional theme
- Hide code in final output (`echo = FALSE`)

**Show-off opportunities (for the overachievers):**
- Add a third variable using size, color, or faceting
- Apply a scale transformation and explain why
- Include data source and analysis attribution in caption

### Part 3: Written Interpretation (5 points)

For each visualization, write 2-3 sentences describing:
1. What the visualization shows
2. Key patterns or relationships you observe
3. Why this information might be important or interesting

## Formatting Requirements

- Set up code chunks with `echo = FALSE` so code is hidden
- Include proper YAML header (already configured)
- Render to PDF format
- Write interpretations in regular text (not code chunks)



## Submission Instructions

Submit **both** files to gradescope:
1. Your `.qmd` source file
2. The rendered `.pdf` file


## Academic Integrity

If you use ChatGPT or other AI tools for assistance, include an appendix detailing:
- Which tools you used
- What specific help you requested
- How you verified and modified the AI-generated content

Remember: The goal is to demonstrate **your** understanding of data visualization principles.


**Now go forth and make some beautiful plots! 📊✨**